R for Clinical Study Reports and Submission

### 2021-10-14 index.Rmd
use sentence case instead of title case for headings; run styler on Rmds

--- 
title: "R for Clinical Study Reports and Submission"
author: ["Yilong Zhang", "Nan Xiao", "Keaven Anderson"]
knit: "bookdown::render_book"
description: >
  This book will teach you how to prepare tables, listings and figures for clinical study report and submit to regulatory agency, 
  the essential part of clinical trial development.

@@

```{r, eval = FALSE, include=FALSE}
r2rtf:::rtf_convert_format(list.files("tlf", pattern = ".rtf", full.names = TRUE), 
                           output_dir = "tlf")
```

@@

## List of Authors and Contributors {-}

```{r, echo = FALSE}
contributors <- c(person("Yilong", "Zhang", email = "yilong.zhang@merck.com", role = "aut"), 
                  person("Nan", "Xiao", role = "aut"), 
                  person("Keaven", "Anderson", role = "aut"),
                  person("Yujie", "Zhao", role = "ctr"), 
                  person("Aiming", "Yang", role = "ctr"),
                  person("Steven", "Haesendonckx", role = "ctr"))
```

>>>>

--- 
title: "R for Clinical Study Reports and Submission"
author: ["Yilong Zhang", "Nan Xiao", "Keaven Anderson"]
knit: "bookdown::render_book"
description: >
  This book will teach you how to prepare tables, listings, and figures for
  clinical study report and submit to regulatory agencies, 
  the essential part of clinical trial development.

@@

```{r, eval = FALSE, include=FALSE}
r2rtf:::rtf_convert_format(
  list.files("tlf", pattern = ".rtf", full.names = TRUE),
  output_dir = "tlf"
)
```

@@

## Authors and contributors {-}

```{r, echo = FALSE}
contributors <- c(
  person("Yilong", "Zhang", email = "yilong.zhang@merck.com", role = "aut"),
  person("Nan", "Xiao", role = "aut"),
  person("Keaven", "Anderson", role = "aut"),
  person("Yujie", "Zhao", role = "ctr"),
  person("Aiming", "Yang", role = "ctr"),
  person("Steven", "Haesendonckx", role = "ctr")
)
```

### 2021-10-14 intro.Rmd
use sentence case instead of title case for headings; run styler on Rmds

- Part 3, "eCTD Submission package"
provides general information in preparing 
submission package related to clinical study report (CSR) in 
electronic Common Technical Document (eCTD). 
In this part, we assume a reader is a project lead of a clinical project 
with experience in R package development and submission.

@@

We share the same philosophy described in 
[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) 
and quote here.

>>>>

- Part 3, "eCTD submission package"
provides general information in preparing 
submission package related to clinical study report (CSR) in 
electronic Common Technical Document (eCTD). 
In this part, we assume a reader is a project lead of a clinical project 
with experience in R package development and submission.

@@

We share the same philosophy described in 
[Section 1.1 of R Packages book](https://r-pkgs.org/intro.html#intro-phil) 
and quote here.

### 2021-10-14 tlf-ae-spec.Rmd
use sentence case instead of title case for headings; run styler on Rmds

```{r}
ana <- adae %>% 
  mutate(AESOC = toTitleCase(tolower(AESOC)), 
         AEDECOD =  toTitleCase(tolower(AEDECOD)))

t1 <-ana %>% 
  group_by(TRTAN, AESOC) %>%
  summarise(n = fmt_num(n_distinct(USUBJID), digits = 0)) %>% 
  mutate(AEDECOD = AESOC, order = 0)

@@

```{r}
t_ae <- bind_rows(t1, t2) %>% 
        pivot_wider(id_cols = c(AESOC, order, AEDECOD), 
                    names_from = TRTAN, 
                    names_prefix = "n_",
                    values_from = n, 
                    values_fill = fmt_num(0, digits = 0)) %>%
        arrange(AESOC, order, AEDECOD) %>% 
        select(AESOC, AEDECOD, starts_with("n"))

@@

```{r}
t_pop <- adsl %>%
  filter(SAFFL == "Y") %>%
  count_by("TRT01AN", "SAFFL",
    var_label = "Participants in population"
  ) %>%
  mutate(
    AESOC = "pop", 
    AEDECOD = var_label) %>%
  select(AESOC,AEDECOD, starts_with("n_"))

@@

```{r}
tbl_ae_spec <- bind_rows(t_pop,
                         data.frame(AESOC = "pop"), 
                         t_ae) %>% 
  mutate(AEDECOD = ifelse(AEDECOD == AESOC , 
                          AEDECOD, paste0("  ", AEDECOD)))

@@

```{r}
tbl_ae_spec %>% 
  rtf_title("Analysis of Participants With Specific Adverse Events", 
            "(Safty Analysis Population)") %>%

    rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(1, 3))
  ) %>%
  rtf_colheader(" | n |  n | n ",
    border_top = "",
    border_bottom = "single",
    col_rel_width = c(3, rep(1, 3))
  ) %>%

  rtf_body(
    col_rel_width = c(1, 3, rep(1, 3)),
    text_justification = c("l", "l", rep("c", 3)),
    text_format = matrix(text_format, nrow = n_row, ncol = n_col), 
    page_by = "AESOC", 
    pageby_row = "first_row") %>%

  rtf_footnote("Every subject is counted a single time for each applicable row and column.") %>%

  rtf_encode() %>%

  write_rtf("tlf/tlf_spec_ae.rtf")

```

>>>>

```{r}
ana <- adae %>%
  mutate(
    AESOC = toTitleCase(tolower(AESOC)),
    AEDECOD = toTitleCase(tolower(AEDECOD))
  )

t1 <- ana %>%
  group_by(TRTAN, AESOC) %>%
  summarise(n = fmt_num(n_distinct(USUBJID), digits = 0)) %>%
  mutate(AEDECOD = AESOC, order = 0)

@@

```{r}
t_ae <- bind_rows(t1, t2) %>%
  pivot_wider(
    id_cols = c(AESOC, order, AEDECOD),
    names_from = TRTAN,
    names_prefix = "n_",
    values_from = n,
    values_fill = fmt_num(0, digits = 0)
  ) %>%
  arrange(AESOC, order, AEDECOD) %>%
  select(AESOC, AEDECOD, starts_with("n"))

@@

```{r}
t_pop <- adsl %>%
  filter(SAFFL == "Y") %>%
  count_by("TRT01AN", "SAFFL",
    var_label = "Participants in population"
  ) %>%
  mutate(
    AESOC = "pop",
    AEDECOD = var_label
  ) %>%
  select(AESOC, AEDECOD, starts_with("n_"))

@@

```{r}
tbl_ae_spec <- bind_rows(
  t_pop,
  data.frame(AESOC = "pop"),
  t_ae
) %>%
  mutate(AEDECOD = ifelse(AEDECOD == AESOC,
    AEDECOD, paste0("  ", AEDECOD)
  ))

@@

```{r}
tbl_ae_spec %>%
  rtf_title(
    "Analysis of Participants With Specific Adverse Events",
    "(Safty Analysis Population)"
  ) %>%
  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(1, 3))
  ) %>%
  rtf_colheader(" | n |  n | n ",
    border_top = "",
    border_bottom = "single",
    col_rel_width = c(3, rep(1, 3))
  ) %>%
  rtf_body(
    col_rel_width = c(1, 3, rep(1, 3)),
    text_justification = c("l", "l", rep("c", 3)),
    text_format = matrix(text_format, nrow = n_row, ncol = n_col),
    page_by = "AESOC",
    pageby_row = "first_row"
  ) %>%
  rtf_footnote("Every subject is counted a single time for each applicable row and column.") %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_spec_ae.rtf")
```

### 2021-10-14 tlf-ae-summary.Rmd
use sentence case instead of title case for headings; run styler on Rmds

# AE Summary {#aesummary}

@@

```{r}
t_pop <- pop %>% 
  mutate(name = "Participants in population", 
         tot = fmt_num(tot, digits = 0)) %>% 
  pivot_wider(id_cols = name, 
              names_from = TRTAN, 
              names_prefix = "n_", 
              values_from = tot)

>>>>

# AE summary {#aesummary}

@@

```{r}
t_pop <- pop %>%
  mutate(
    name = "Participants in population",
    tot = fmt_num(tot, digits = 0)
  ) %>%
  pivot_wider(
    id_cols = name,
    names_from = TRTAN,
    names_prefix = "n_",
    values_from = tot
  )

### 2021-10-14 tlf-baseline.Rmd
use sentence case instead of title case for headings; run styler on Rmds

# Baseline Characteristics

@@

```{r}
ana <- adsl %>% 
  mutate(SEX = factor(SEX, c("F", "M"), c("Female", "Male")), 
         RACE = toTitleCase(tolower(RACE)))

@@

```{r}
tbl_base <- tbl %>% 
              as.data.frame() %>% 
              as_tibble() %>%
              mutate( across(everything(), 
                      ~ str_replace_all(.x, intToUtf8(160), " ") ) 
                    )

@@

tbl_base[-1, ] %>% 
      rtf_title("Participant Baseline Characteristics", 
                "(All Participants Randomized)") %>%
      rtf_colheader(colheader1, 
                    col_rel_width = rel_width) %>% 
      rtf_colheader(colheader2, 
                    border_top = "",
                    col_rel_width = rel_width) %>%
      rtf_body(col_rel_width = rel_width,
               text_justification = c("l", rep("c", 4)), 
               text_indent_first = -240,
               text_indent_left = 180) %>% 
      rtf_encode() %>% 
      write_rtf("tlf/tlf_base.rtf")
```

>>>>

# Baseline characteristics

@@

```{r}
ana <- adsl %>%
  mutate(
    SEX = factor(SEX, c("F", "M"), c("Female", "Male")),
    RACE = toTitleCase(tolower(RACE))
  )

@@

```{r}
tbl_base <- tbl %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate(across(
    everything(),
    ~ str_replace_all(.x, intToUtf8(160), " ")
  ))

@@

tbl_base[-1, ] %>%
  rtf_title(
    "Participant Baseline Characteristics",
    "(All Participants Randomized)"
  ) %>%
  rtf_colheader(colheader1,
    col_rel_width = rel_width
  ) %>%
  rtf_colheader(colheader2,
    border_top = "",
    col_rel_width = rel_width
  ) %>%
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 4)),
    text_indent_first = -240,
    text_indent_left = 180
  ) %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_base.rtf")
```

### 2021-10-14 tlf-disposition.Rmd
use sentence case instead of title case for headings; run styler on Rmds

```{r}
tbl_disp %>%

  # Table title
  rtf_title("Disposition of Participants") %>%

  # First row of column header
  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(2, 3))
  ) %>%

  # Second row of column header
  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    border_top = c("", rep("single", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%

  # Table body
  rtf_body(
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    text_justification = c("l", rep("c", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%

  # Encoding RTF syntax
  rtf_encode() %>%

  # Save to a file
  write_rtf("tlf/tbl_disp.rtf")
```

>>>>

```{r}
tbl_disp %>%
  # Table title
  rtf_title("Disposition of Participants") %>%
  # First row of column header
  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(2, 3))
  ) %>%
  # Second row of column header
  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    border_top = c("", rep("single", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%
  # Table body
  rtf_body(
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    text_justification = c("l", rep("c", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%
  # Encoding RTF syntax
  rtf_encode() %>%
  # Save to a file
  write_rtf("tlf/tbl_disp.rtf")
```

### 2021-10-14 tlf-efficacy.Rmd
use sentence case instead of title case for headings; run styler on Rmds

## Analysis Dataset

@@

```{r}
ana %>%
  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%  
  head(4)
```
## Helper Functions

@@

## Summary of Observed Data

@@

## Missing Data Imputation

@@

```{r}
t1_rtf <- t1 %>% 
  data.frame() %>%
  rtf_title(c(
    "ANCOVA of Change from Baseline Glucose (mmol/L) at Week 24",
    "LOCF",
    "Efficacly Analysis Population"
  )) %>%
  rtf_colheader("| Baseline | Week 24 | Change from Baseline",
    col_rel_width = c(2.5, 2, 2, 4)
  ) %>%
  rtf_colheader(paste(
    "Treatment |",
    paste0(rep("N | Mean (SD) | ", 3), collapse = ""),
    "LS Mean (95% CI){^a}"
  ),
  col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_body(text_justification = c("l", rep("c", 7)), 
           col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)) %>%
  rtf_footnote(c(
    "{^a}Based on an ANCOVA model after adjusting baseline value. LOCF approach is used to impute missing values.",
    "ANCOVA = Analysis of Covariance, LOCF = Last Observation Carried Forward",
    "CI = Confidence Interval, LS = Least Squares, SD = Standard Deviation"
  ))

@@

```{r}
t2_rtf <- t2 %>% 
  data.frame() %>%
  rtf_colheader("Pairwise Comparison | Difference in LS Mean (95% CI){^a} | p-Value",
    col_rel_width = c(4.5, 4, 2)
  ) %>%
  rtf_body(text_justification = c("l", "c", "c"), 
           col_rel_width = c(4.5, 4, 2))

>>>>

## Analysis dataset

@@

```{r}
ana %>%
  select(USUBJID, TRTPN, AVISIT, AVAL, BASE, CHG) %>%
  head(4)
```
## Helper functions

@@

## Summary of observed data

@@

## Missing data imputation

@@

```{r}
t1_rtf <- t1 %>%
  data.frame() %>%
  rtf_title(c(
    "ANCOVA of Change from Baseline Glucose (mmol/L) at Week 24",
    "LOCF",
    "Efficacly Analysis Population"
  )) %>%
  rtf_colheader("| Baseline | Week 24 | Change from Baseline",
    col_rel_width = c(2.5, 2, 2, 4)
  ) %>%
  rtf_colheader(paste(
    "Treatment |",
    paste0(rep("N | Mean (SD) | ", 3), collapse = ""),
    "LS Mean (95% CI){^a}"
  ),
  col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_body(
    text_justification = c("l", rep("c", 7)),
    col_rel_width = c(2.5, rep(c(0.5, 1.5), 3), 2)
  ) %>%
  rtf_footnote(c(
    "{^a}Based on an ANCOVA model after adjusting baseline value. LOCF approach is used to impute missing values.",
    "ANCOVA = Analysis of Covariance, LOCF = Last Observation Carried Forward",
    "CI = Confidence Interval, LS = Least Squares, SD = Standard Deviation"
  ))

@@

```{r}
t2_rtf <- t2 %>%
  data.frame() %>%
  rtf_colheader("Pairwise Comparison | Difference in LS Mean (95% CI){^a} | p-Value",
    col_rel_width = c(4.5, 4, 2)
  ) %>%
  rtf_body(
    text_justification = c("l", "c", "c"),
    col_rel_width = c(4.5, 4, 2)
  )

### 2021-10-14 tlf-overview.Rmd
use sentence case instead of title case for headings; run styler on Rmds

## Structure and Content

@@

- Figure out table layout.

- Split the layout into small tasks in the form of a computer program.

- Execute the program.

@@

```{r}
head(tbl) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%

  # define relative width
  rtf_encode() %>%
  write_rtf("tlf/intro-ae2.rtf")
```

@@

```{r}
head(tbl) %>%

  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    col_rel_width = c(3, 2, 2, 2)
  ) %>%

  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%

  rtf_encode() %>%

  write_rtf("tlf/intro-ae3.rtf")
```

@@

```{r}
head(tbl) %>%

  rtf_body(text_justification = c("l", "c", "c", "c")) %>%

  rtf_encode() %>%

  write_rtf("tlf/intro-ae5.rtf")
```

@@

```{r}
head(tbl) %>%

  rtf_colheader(
    colheader = " | Treatment",
    col_rel_width = c(3, 6)
  ) %>%

  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    border_top = c("", "single", "single", "single"),
    col_rel_width = c(3, 2, 2, 2)
  ) %>%

  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%

  rtf_encode() %>%

  write_rtf("tlf/intro-ae7.rtf")
```

>>>>

## Structure and content

@@

- Figure out table layout. 
- Split the layout into small tasks in the form of a computer program.
- Execute the program.

@@

```{r}
head(tbl) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  # define relative width
  rtf_encode() %>%
  write_rtf("tlf/intro-ae2.rtf")
```

@@

```{r}
head(tbl) %>%
  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae3.rtf")
```

@@

```{r}
head(tbl) %>%
  rtf_body(text_justification = c("l", "c", "c", "c")) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae5.rtf")
```

@@

```{r}
head(tbl) %>%
  rtf_colheader(
    colheader = " | Treatment",
    col_rel_width = c(3, 6)
  ) %>%
  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    border_top = c("", "single", "single", "single"),
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae7.rtf")
```

### 2021-10-14 tlf-population.Rmd
use sentence case instead of title case for headings; run styler on Rmds

# Analysis Population {#population}

@@

## Helper Functions

@@

```{r}
count_by(adsl, "TRT01PN", "EFFFL") %>% 
  select(- ends_with(c("_54", "_81")))
```

## Analysis Code

@@

pop <- count_by(adsl, "TRT01PN", "RANDFL", 
                var_label = "Participants in Population") %>%
  select(var_label, starts_with("n_"))
```

>>>>

# Analysis population {#population}

@@

## Helper functions

@@

```{r}
count_by(adsl, "TRT01PN", "EFFFL") %>%
  select(-ends_with(c("_54", "_81")))
```

## Analysis code

@@

pop <- count_by(adsl, "TRT01PN", "RANDFL",
  var_label = "Participants in Population"
) %>%
  select(var_label, starts_with("n_"))
```
